### 2023-10-04 chapters/paper_modelagnostic.tex
Work on font and fix typos

The ability to recognise when data are anomalous, i.e. if they originate from a distribution different from that of the training data, is a necessary property for machine learning models for safe and reliable applications in the real world. Historically, \textcite{bishop_novelty_1994} proposed to use a one-sided threshold on the log-likelihoods of a learned model as a decision rule to identify outliers in a dataset. However, recently, \textcite{nalisnick_deep_2019,hendrycks_deep_2019} showed that state-of-the-art deep generative models (DGMs) failed in this task, assigning higher a likelihood to out-of-distribution (OOD) data than in-distribution data. Most of the recent works focused on proposing new test statistics to alleviate the problem of using the plain likelihood, see \cref{sec_modelagnostic:related_works} for details.

We believe that OOD detection should be formulated as statistical hypothesis testing \parencite{nalisnick_detecting_2019, ahmadian_likelihoodfree_2021, haroush_statistical_2021}. Since the power of a single test depends on the out-distribution \parencite{zhang_understanding_2021}, we propose to approach this problem by using a combination of multiple statistical tests. %While this does not solve completely the problem, because the power of the combined tests will also depend on the out-distribution, we think that we can get better performance than using a single-statistics especially in situations where one statistic fails. 
While the power of the combined test also depends on the out-distribution, we hypothesise that the combined test empirically will perform better, especially in situations where one of the statistics fails.
%
%Furthermore, the use of the statistical testing framework, allows us to correct for the multiple comparisons problem when identifying outliers in a dataset by controlling the number of type I errors through the false discovery rate (FDR).
Furthermore, the use of the statistical testing framework has several advantages. Since we obtain a $p$-value, it is more natural deciding on a threshold as this corresponds to the significance level. In addition to that, it also allow us to correct for the multiple comparisons problem when identifying outliers in a dataset by controlling the number of Type I errors through the false discovery rate (FDR).

@@

\paragraph{The typicality kernel} A very simple approach of embedding the data using $p_\theta$ is to choose $\Phi_\textup{Typical}(x) = \log p_\theta (x)$. Then, MMD is exactly equivalent to the \emph{typicality test statistic} of \textcite{nalisnick_detecting_2019}, although this connection was not explicitly stated  by \textcite{nalisnick_detecting_2019}. Because of this, we call the kernel $k(x,y) = \log p_\theta (x) \cdot \log p_\theta (y)$ the \emph{typicality kernel}.  %We could also call this "new" kernel the "likelihood kernel".
%While $\Phi_\textup{Typical}$ is not as well motivated as the Fisher kernel, we found that it generally gives good results. 
While $\Phi_\textup{Typical}$ is not as well motivated as a kernel as $\Phi_\textup{Fisher}$, the concepts of typicality and typical set can be used to explain unintuitive behaviours of probability distributions in high-dimensional space as highlighted by \textcite{nalisnick_deep_2019}. We also found that using this kernel generally gives good results for OOD tasks. An interesting analysis that we d not consider in this paper would be to study the properties of this kernel.

@@

%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%  SECTION 3   %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Combining different test statistics}
\label{sec_modelagnostic:combination_p_values}
For single-sample OOD detection, \textcite{zhang_understanding_2021} proved that there is not a single statistic that is constantly better compared to all the possible alternatives of interest. For this reason, we believe that using a combination of different test statistics should lead to an overall better OOD detection in settings where a single statistic might fail. Assume we compute $k$ different test statistics $T_1,\dots, T_k$, each testing $\mathcal{H}_0$ against $\mathcal{H}$ as defined in \cref{sec_modelagnostic:parametric,sec_modelagnostic:MMD}. The goal is to combine these different tests into a single statistical test that ideally will perform better than the initial single tests. However, different tests can have different magnitudes and they can differ also in the direction of out-of-distribution detection, i.e.\@ for some statistics having a higher values is associated with being OOD, while for other smaller values are OOD. This makes a combination non-trivial.

\textcite{morningstar_density_2021} proposed the density of states estimator (DoSE) to overcome this problem. They only focused on the single-sample detection task, i.e.\@ $n=1$ following our problem formulation. Their idea is to fit different nonparametric density estimators, such as a kernel-density estimator (KDE) or a one-class support vector machines (SVM), for each different statistic $T_1,\dots, T_k$ by using the values computed on the training set examples. For a single test example, $\tilde{x}_1$, they first compute $T_1,\dots, T_k$ and then combine those statistics by summing the different KDEs log-density. %This approach can be used when combining both one-sided and two-sided test statistic. 
While this approach can be used for any type of statistic, and thus is more general, it uses less prior information. Indeed, if we use only statistics that are truly one-sided, then we assume that a method that leverages the true nature of the statistics should work better. In addition to that, fitting a KDE introduces an additional hyperparameter.

@@

A possible third option would be to estimate the diagonal of $I(\theta)$ using samples from the model. However, for autoregressive models as the PixelCNN, sampling is a sequential procedure and therefore it is computationally expensive to generate many samples when the input-space is high-dimensional. For this reason, we do not consider it in this work. More complex and precise approximations of the FIM exists, such as the Kronecker-factored Approximate Curvature (K-FAC, \citealp{martens_optimizing_2015}), but these are not defined for all types of layers used by state-of-the-art models.

\paragraph{On the difficulty of computing per-example gradients} Both the diagonal approximation of the FIM and the computation of the MMD with Fisher kernel of \cref{eq_modelagnostic:mmd_fisher} require the gradient computation for all training and test examples. This is known as a costly procedure. For example, if we have to compute the gradient for $N$ examples using a simple fully connected network with $l$ layers of size $p$, the naive procedure of using a batch-size of dimension 1 is $\mathcal{O}(Nlp^2)$ \parencite{goodfellow_efficient_2015}. While more efficient per-example gradient computations were proposed \parencite{goodfellow_efficient_2015, rochette_efficient_2019}, these techniques can only be applied on simple fully connected or convolutional networks. While for this paper we relied on the naive solution of looping through every example one at the time, a more efficient solution is provided by the BackPACK library \parencite{dangel_backpack_2020} which allows to compute the gradient with respect each sample in a minibatch.

@@

The typicality set hypothesis was introduced by \textcite{nalisnick_detecting_2019} as a possible explanation for the DGMs assigning higher likelihood to OOD data. The typicality set is the subset of the model full support where the model samples from and this does not intersect with the region of higher likelihood. While the typicality test was introduced for batch-OOD detection, \textcite{morningstar_density_2021} shows that it also works well in the single-sample case. This is also confirmed by our own experiments.

@@

As mentioned in the introduction, we frame the OOD detection problem in terms of statistical tests problem. Recently, \textcite{haroush_statistical_2021} showed that adopting hypothesis testing at the layer and channel level of a neural network can be used for OOD detection in the discriminative setting. They used both Fisher's method and Simes' method to combine class-conditional $p$-values computed for each convolutional and dense layer of a deep neural network. We focus on the unsupervised setting using DGMs and use hypothesis testing on statistics that can be computed on all differentiable DGM. As already explained in section \cref{sec_modelagnostic:combination_p_values}, \textcite{morningstar_density_2021} considered the combination of different statistics for OOD detection. The main difference with their approach is that we propose statistics that can be applied to any differentiable generative model and combine them by using Fisher's method, which takes advantage of using only one-sided independent statistics. Concurrently, \textcite{choi_robust_2021} derived the score statistic by starting from the likelihood ratio statistic and applying a Laplace approximation. They computed the score statistic only for certain layers of the model and for a specific example, the OOD score is given by the infinity norm of these different layer scores after a ReLU operation. Our procedure differs both in the derivation of the score statistic and its usage since we compute the score statistic for the entire model.

\begin{sidewaysfigure}[tbp]
    \includegraphics[width=\textwidth]{graphics/paper_modelagnostic/pvalues_comb_no_drop_correct_font.pdf}
    \caption[Distribution of the p-values of the typicality test, score statistic, Fisher's method and a combination of all.]{\emph{First plot}: $p$-values of the typicality test on the two test sets. We can see that under $\mathcal{H}_0$, they should be uniformly distributed. \emph{Second plot}: $p$-values of the score statistic. \emph{Third plot}: values obtained by the Fisher's method. In red, we plot the density function of a $\chi^2$-distribution with 4 degrees. This shows that the statistics are independent. \emph{Fourth plot}: $p$-values obtained of the combination. These plots refer to a PixelCNN++ trained on FashionMNIST without dropout.}
    \label{fig_modelagnostic:comb_p_values}
\end{sidewaysfigure}

@@

\paragraph{Two-sample OOD}
As \textcite{nalisnick_detecting_2019}, we consider how these test statistics change when performing two-sample OOD detection. Results are summarized in \cref{tab_modelagnostic:two_sample_results}. As shown by \textcite{nalisnick_detecting_2019}, the typicality improves but also the score statistic gets better if we consider more samples. Combining those leads to an improvement of performance in terms of AUROC with almost all the models. When training on FashionMNIST, the model can almost perfectly distinguish between the in-distribution test set and the OOD test set. While the performance improves for the two \textit{far}-OOD task, we have that the improvement is slightly less evident in the \textit{near}-OOD task of CIFAR10 vs CIFAR100.

@@

\subsection{Practical OOD detection with FDR control}
One of the advantages of framing the problem as multiple testing is that we have a well-defined procedure to decide on which hypotheses to reject while controlling the False Discovery Rate (FDR, \citealp{benjamini_controlling_1995}). Imagine we are interested in finding the outliers from the dataset given by the combination of the two test-sets but we do not want to discard too many inliers, then we can use the Benjamini-Hochberg (BH) procedure \parencite{benjamini_controlling_1995} to decide a threshold and reject all hypothesis below that threshold. For a specific significance level $\alpha$, the procedure guarantees that the FDR stays below that level. Therefore, we can guarantee that the rate of inliers that are classified as outliers is less than the chosen $\alpha$.

@@

We leverage the fact that when the null hypothesis is true and the $p$-values are independent, then the scores obtained by combining $k$ different statistics are $\chi^2_{2k}$ distributed to compute the $p$-values. Alternatively, th e procedure can be also applied to the $p$-values of a single test-statistic.
%
Usually, it is better to use a FDR control when it is actually possible to make few false discoveries, i.e.\@ when we have a strong statistic. Therefore, we expect the procedure to work well when the AUROC is good, for examples on models trained on FashionMNIST.

As can be seen in \cref{fig_modelagnostic:type1}, we have that the Type I ratio line stays below the identity line, meaning that the BH correction is working. When deciding for a specific threshold $\alpha$, we usually have to trade-off between Type I and Type II error and in most cases the threshold to choose depends on the application domain. Ideally, we would like to have a low Type I and a low Type II error rate, meaning that we are not considering a lot of in-distribution examples as OOD and at the same time considering a lot of outliers as in-distribution. \cref{fig_modelagnostic:type1} shows that we can achieve this for low values of $\alpha$. When training on CIFAR, instead, we are able to control the FDR only from a certain significance level (see \cref{appendix_modelagnostic:BH_cifar}). This is expected given that the AUROC is not as good as when testing on MNIST.

@@

When considering only one-sided independent statistics, we showed that the Fisher's method tends to works better than combine them by summing the log-density of a KDE. We also noticed that the score statistic tends to perform a bit worse when the number of parameters of the models increases, i.e.\@ in the context of natural images. One possible reason can be that in this setting the diagonal approximation is not good, and therefore one could consider different approximations, such as K-FAC.

@@

\section*{Acknowledgements}
Federico Bergamin and Pierre-Alexandre Mattei contributed eqaully to this paper, which is indicated by the asterisk (*) in the author list. The work was supported by the Innovation Fund Denmark (0175-00014B and 0153-00167B), the Independent Research Fund Denmark (9131-00082B) and the Novo Nordisk Foundation (NNF20OC0062606 and NNF20OC0065611). Furthermore, it was supported by the French government, through the 3IA Côte d’Azur Investments in the Future project managed by the National Research Agency (ANR) with the reference number ANR-19-P3IA-0002.

>>>>

The ability to recognize when data are anomalous, i.e. if they originate from a distribution different from that of the training data, is a necessary property for machine learning models for safe and reliable applications in the real world. Historically, \textcite{bishop_novelty_1994} proposed to use a one-sided threshold on the log-likelihoods of a learned model as a decision rule to identify outliers in a dataset. However, recently, \textcite{nalisnick_deep_2019,hendrycks_deep_2019} showed that state-of-the-art deep generative models (DGMs) failed in this task, assigning higher a likelihood to out-of-distribution (OOD) data than in-distribution data. Most of the recent works focused on proposing new test statistics to alleviate the problem of using the plain likelihood, see \cref{sec_modelagnostic:related_works} for details.

We believe that OOD detection should be formulated as statistical hypothesis testing \parencite{nalisnick_detecting_2019, ahmadian_likelihoodfree_2021, haroush_statistical_2021}. Since the power of a single test depends on the out-distribution \parencite{zhang_understanding_2021}, we propose to approach this problem by using a combination of multiple statistical tests. %While this does not solve completely the problem, because the power of the combined tests will also depend on the out-distribution, we think that we can get better performance than using a single-statistics especially in situations where one statistic fails. 
While the power of the combined test also depends on the out-distribution, we hypothesize that the combined test empirically will perform better, especially in situations where one of the statistics fails.
%
%Furthermore, the use of the statistical testing framework, allows us to correct for the multiple comparisons problem when identifying outliers in a dataset by controlling the number of type I errors through the false discovery rate (FDR).
Furthermore, the use of the statistical testing framework has several advantages. Since we obtain a $p$-value, it is more natural deciding on a threshold as this corresponds to the significance level. In addition to that, it also allows us to correct for the multiple comparisons problem when identifying outliers in a dataset by controlling the number of Type I errors through the false discovery rate (FDR).

@@

\paragraph{The typicality kernel} A very simple approach of embedding the data using $p_\theta$ is to choose $\Phi_\textup{Typical}(x) = \log p_\theta (x)$. Then, MMD is exactly equivalent to the \emph{typicality test statistic} of \textcite{nalisnick_detecting_2019}, although this connection was not explicitly stated  by \textcite{nalisnick_detecting_2019}. Because of this, we call the kernel $k(x,y) = \log p_\theta (x) \cdot \log p_\theta (y)$ the \emph{typicality kernel}.  %We could also call this "new" kernel the "likelihood kernel".
%While $\Phi_\textup{Typical}$ is not as well motivated as the Fisher kernel, we found that it generally gives good results. 
While $\Phi_\textup{Typical}$ is not as well motivated as a kernel as $\Phi_\textup{Fisher}$, the concepts of typicality and typical set can be used to explain unintuitive behaviours of probability distributions in high-dimensional space as highlighted by \textcite{nalisnick_deep_2019}. We also found that using this kernel generally gives good results for OOD tasks. An interesting analysis that we'd not consider in this paper would be to study the properties of this kernel.

@@

%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%  SECTION 3   %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Combining different test statistics}
\label{sec_modelagnostic:combination_p_values}
For single-sample OOD detection, \textcite{zhang_understanding_2021} proved that there is not a single statistic that is constantly better compared to all the possible alternatives of interest. For this reason, we believe that using a combination of different test statistics should lead to an overall better OOD detection in settings where a single statistic might fail. Assume we compute $k$ different test statistics $T_1,\dots, T_k$, each testing $\mathcal{H}_0$ against $\mathcal{H}$ as defined in \cref{sec_modelagnostic:parametric,sec_modelagnostic:MMD}. The goal is to combine these different tests into a single statistical test that ideally will perform better than the initial single tests. However, different tests can have different magnitudes, and they can differ also in the direction of out-of-distribution detection, i.e.\@ for some statistics having a higher value is associated with being OOD, while for other smaller values are OOD. This makes a combination non-trivial.

\textcite{morningstar_density_2021} proposed the density of states estimator (DoSE) to overcome this problem. They only focused on the single-sample detection task, i.e.\@ $n=1$ following our problem formulation. Their idea is to fit different non-parametric density estimators, such as a kernel-density estimator (KDE) or a one-class support vector machines (SVM), for each different statistic $T_1,\dots, T_k$ by using the values computed on the training set examples. For a single test example, $\tilde{x}_1$, they first compute $T_1,\dots, T_k$ and then combine those statistics by summing the different KDEs log-density. %This approach can be used when combining both one-sided and two-sided test statistic. 
While this approach can be used for any type of statistic, and thus is more general, it uses less prior information. Indeed, if we use only statistics that are truly one-sided, then we assume that a method that leverages the true nature of the statistics should work better. In addition to that, fitting a KDE introduces an additional hyperparameter.

@@

A possible third option would be to estimate the diagonal of $I(\theta)$ using samples from the model. However, for autoregressive models as the PixelCNN, sampling is a sequential procedure, and therefore it is computationally expensive to generate many samples when the input-space is high-dimensional. For this reason, we do not consider it in this work. More complex and precise approximations of the FIM exists, such as the Kronecker-factored Approximate Curvature (K-FAC, \citealp{martens_optimizing_2015}), but these are not defined for all types of layers used by state-of-the-art models.

\paragraph{On the difficulty of computing per-example gradients} Both the diagonal approximation of the FIM and the computation of the MMD with Fisher kernel of \cref{eq_modelagnostic:mmd_fisher} require the gradient computation for all training and test examples. This is known as a costly procedure. For example, if we have to compute the gradient for $N$ examples using a simple fully connected network with $l$ layers of size $p$, the naive procedure of using a batch-size of dimension 1 is $\mathcal{O}(Nlp^2)$ \parencite{goodfellow_efficient_2015}. While more efficient per-example gradient computations were proposed \parencite{goodfellow_efficient_2015, rochette_efficient_2019}, these techniques can only be applied on simple fully connected or convolutional networks. While for this paper we relied on the naive solution of looping through every example one at the time, a more efficient solution is provided by the BackPACK library \parencite{dangel_backpack_2020} which allows to compute the gradient with respect each sample in a mini-batch.

@@

The typicality set hypothesis was introduced by \textcite{nalisnick_detecting_2019} as a possible explanation for the DGMs assigning higher likelihood to OOD data. The typicality set is the subset of the model full support where the model samples from and this does not intersect with the region of higher likelihood. While the typicality test was introduced for batch OOD detection, \textcite{morningstar_density_2021} shows that it also works well in the single-sample case. This is also confirmed by our own experiments.

@@

As mentioned in the introduction, we frame the OOD detection problem in terms of a statistical test problem. Recently, \textcite{haroush_statistical_2021} showed that adopting hypothesis testing at the layer and channel level of a neural network can be used for OOD detection in the discriminative setting. They used both Fisher's method and Simes' method to combine class-conditional $p$-values computed for each convolutional and dense layer of a deep neural network. We focus on the unsupervised setting using DGMs and use hypothesis testing on statistics that can be computed on all differentiable DGM. As already explained in section \cref{sec_modelagnostic:combination_p_values}, \textcite{morningstar_density_2021} considered the combination of different statistics for OOD detection. The main difference with their approach is that we propose statistics that can be applied to any differentiable generative model and combine them by using Fisher's method, which takes advantage of using only one-sided independent statistics. Concurrently, \textcite{choi_robust_2021} derived the score statistic by starting from the likelihood ratio statistic and applying a Laplace approximation. They computed the score statistic only for certain layers of the model and for a specific example, the OOD score is given by the infinity norm of these different layer scores after a ReLU operation. Our procedure differs both in the derivation of the score statistic and its usage since we compute the score statistic for the entire model.

\begin{sidewaysfigure}[tbp]
    \includegraphics[width=\textwidth]{graphics/paper_modelagnostic/pvalues_comb_no_drop_correct_font.pdf}
    \caption[Distribution of the p-values of the typicality test, score statistic, Fisher's method and a combination of all.]{ \emph{First plot}: $p$-values of the typicality test on the two test sets. We can see that under $\mathcal{H}_0$, they should be uniformly distributed. \emph{Second plot}: $p$-values of the score statistic. \emph{Third plot}: values obtained by the Fisher's method. In red, we plot the density function of a $\chi^2$-distribution with 4 degrees. This shows that the statistics are independent. \emph{Fourth plot}: $p$-values obtained of the combination. These plots refer to a PixelCNN++ trained on FashionMNIST without dropout.}
    \label{fig_modelagnostic:comb_p_values}
\end{sidewaysfigure}

@@

\paragraph{Two-sample OOD}
As \textcite{nalisnick_detecting_2019}, we consider how these test statistics change when performing two-sample OOD detection. Results are summarized in \cref{tab_modelagnostic:two_sample_results}. As shown by \textcite{nalisnick_detecting_2019}, the typicality improves, but also the score statistic gets better if we consider more samples. Combining those leads to an improvement of performance in terms of AUROC with almost all the models. When training on FashionMNIST, the model can almost perfectly distinguish between the in-distribution test set and the OOD test set. While the performance improves for the two \textit{far}-OOD task, we have that the improvement is slightly less evident in the \textit{near}-OOD task of CIFAR10 vs CIFAR100.

@@

\subsection{Practical OOD detection with FDR control}
One of the advantages of framing the problem as multiple testing is that we have a well-defined procedure to decide on which hypotheses to reject while controlling the False Discovery Rate (FDR, \citealp{benjamini_controlling_1995}). Imagine we are interested in finding the outliers from the dataset given by the combination of the two test-sets, but we do not want to discard too many inliers, then we can use the Benjamini-Hochberg (BH) procedure \parencite{benjamini_controlling_1995} to decide a threshold and reject all hypothesis below that threshold. For a specific significance level $\alpha$, the procedure guarantees that the FDR stays below that level. Therefore, we can guarantee that the rate of inliers that are classified as outliers is less than the chosen $\alpha$.

@@

We leverage the fact that when the null hypothesis is true and the $p$-values are independent, then the scores obtained by combining $k$ different statistics are $\chi^2_{2k}$ distributed to compute the $p$-values. Alternatively, the procedure can be also applied to the $p$-values of a single test-statistic.
%
Usually, it is better to use an FDR control when it is actually possible to make few false discoveries, i.e.\@ when we have a strong statistic. Therefore, we expect the procedure to work well when the AUROC is good, for examples on models trained on FashionMNIST.

As can be seen in \cref{fig_modelagnostic:type1}, we have that the Type I ratio line stays below the identity line, meaning that the BH correction is working. When deciding for a specific threshold $\alpha$, we usually have to trade off between Type I and Type II error and in most cases the threshold to choose depends on the application domain. Ideally, we would like to have a low Type I and a low Type II error rate, meaning that we are not considering a lot of in-distribution examples as OOD and at the same time considering a lot of outliers as in-distribution. \cref{fig_modelagnostic:type1} shows that we can achieve this for low values of $\alpha$. When training on CIFAR, instead, we are able to control the FDR only from a certain significance level (see \cref{appendix_modelagnostic:BH_cifar}). This is expected given that the AUROC is not as good as when testing on MNIST.

@@

When considering only one-sided independent statistics, we showed that the Fisher's method tends to work better than combine them by summing the log-density of a KDE. We also noticed that the score statistic tends to perform a bit worse when the number of parameters of the models increases, i.e.\@ in the context of natural images. One possible reason can be that in this setting the diagonal approximation is not good, and therefore one could consider different approximations, such as K-FAC.

@@

\section*{Acknowledgements}
Federico Bergamin and Pierre-Alexandre Mattei contributed equally to this paper, which is indicated by the asterisk (*) in the author list. The work was supported by the Innovation Fund Denmark (0175-00014B and 0153-00167B), the Independent Research Fund Denmark (9131-00082B) and the Novo Nordisk Foundation (NNF20OC0062606 and NNF20OC0065611). Furthermore, it was supported by the French government, through the 3IA Côte d'Azur Investments in the Future project managed by the National Research Agency (ANR) with the reference number ANR-19-P3IA-0002.
